An Improved Approximate-Bayesian Model-choice Method for Estimating Shared Evolutionary History

### 2014-02-21 manuscripts/dpp-msbayes/dpp-msbayes-main.tex
More editing of the DPP ms.

\begin{linenomath}
Each \divTime{} within \divTimeVector is \iid as $\divTime{} \sim
Gamma(\gammaShape{\divTime{}}, \gammaScale{\divTime{}})$.
Thus, given the number of unique divergence-time classes in
\divTimeIndexVector, this determines the probability of prior term
$p(\divTimeVector \given \divTimeIndexVector)$.
The divergence times are in coalescent units relative to the size of
a constant reference population, which I denote \myTheta{C}, that is equal to
the expectation of the prior on the size of the descendant populations
\begin{equation}
    % \myTheta{C} = \mathbb{E}(\dThetas),
    \myTheta{C} = \mathbb{E}(\descendantTheta{}{}),
    \label{eq:thetaC}
\end{equation}
% where \popIndex is the descendant-population index which can be either 1 or 2.
Given the size of the descendant populations are \iid as
$\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$ and
$\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
this becomes
\begin{equation}
    \myTheta{C} = \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}.
    \label{eq:thetaCGamma}
\end{equation}
The \divTime{} parameters are in units of $\myTheta{C}/\mutationRate$
generations, which I denote as \globalcoalunit generations.
Thus, each \divTime{} within \divTimeVector is proportional to time and can be
converted to the number of generations of the reference population, which I
denote \divTime{G_C}, by assuming a mutation rate and multiplying by the
effective size of the reference population
\begin{equation}
    \divTime{G_C} = \divTime \times \frac{\myTheta{C}}{\mutationRate} = \divTime
    \times \frac{
    \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}}
    {\mutationRate}.
    \label{eq:divTimeGenerations}
\end{equation}
Thus, for each of the divergence times in \divTimeVector to be on the same
scale, the relative mutation rates among the pairs of populations are assumed
to be known and fixed according to the user-provided values in
\mutationRateScalarConstantVector.
\end{linenomath}

@@

The second prior is based on the Dirichlet-process, which is a
stochastic process that groups elements into an unknown number of discrete
parameter classes \citep{Ferguson1973,Antoniak1974}.
The process is controlled by the concentration parameter \concentrationParam,
which determines how clustered the process will be.
I use the Dirichlet process to place a prior over all possible ordered
partitions of \npairs{} elements, which I denote as $\divTimeIndexVector \sim
\priorDPP{}$.
I take a hierarchical approach and use a prior probability
distribution (i.e., hyperprior) for \concentrationParam.
More specifically, I use a gamma-distributed prior $\concentrationParam \sim
Gamma(\gammaShape{\concentrationParam}, \gammaScale{\concentrationParam})$,
where \gammaShape{\concentrationParam} and \gammaScale{\concentrationParam} are
specified by the investigator.

@@

\subsection{Differences between my model and the original \texttt{msBayes} model}
\subsubsection{The prior on divergence models}
One of the key differences between my model and that of \msb \citep{Huang2011}
is the prior distribution on divergence models.
As discussed in \citet{Oaks2012}, in \msb the prior used for
\divTimeIndexVector is a combination of a discrete uniform prior over the
possible number of divergence events \divTimeNum from 1 to \npairs{} with a
multinomial distribution on the number of times each index of \divTimeVector
appears in \divTimeIndexVector, with the constraint that all \divTime{}
parameters are represented at least once (see Equation 2 of \citet{Oaks2012}).
I denote this prior used by \msb as $\divTimeIndexVector \sim \priorOld$.
\citet{Oaks2012} discuss how placing a uniform prior over the number of
divergence parameters (denoted \divTimeNum here, and as \numt{} in
\citet{Huang2011}) imposes an U-shaped prior over discrete divergence
models (\divTimeIndexVector; see
Figure 5B of \citet{Oaks2012}).
To avoid this, I place priors directly on the sample space of divergence
models, thus eliminating the parameter \numt{} from the model.
I introduce two priors on divergence models:
(1) a prior that is uniform over all unordered divergence models, and
(2) a Dirichlet-process prior on all ordered divergence models.
The latter provides an investigator with a great deal of flexibility in
expressing their prior beliefs about models of divergence.

@@

\subsubsection{Flexibility in parameterizing the model}
In my implementation, I provide the ability to control the richness of the
model.
For the \myTheta{} parameters, the model can be fully generalized to
allow each population pair to have three parameters:
\ancestralTheta{}, \descendantTheta{1}{}, and \descendantTheta{2}{}.
Furthermore, any model of \myTheta{} parameters nested within this
general model can also be specified, including the most restricted model
where the ancestral and descendant populations of each pair share
a single \myTheta{} parameter.

@@

Overall, my implementation allows an investigator to specify a model that has
as many as seven parameters per population pair
(\ancestralTheta{}, \descendantTheta{1}{}, \descendantTheta{2}{},
\bottleTime{}, \bottleScalar{1}{}, \bottleScalar{2}{}, and
\migrationRate{})
or as few as one parameter per pair
(\myTheta{}),
in addition to the $\popSampleSize{i}{} - 1$ coalescence-time parameters.

\subsubsection{Time scale}
As described above, in my model divergence times are in units of
$\myTheta{C}/\mutationRate$ generations, where \myTheta{C} is the expectation
of the prior on descendant-population size.
As described by \citet{Oaks2012}, in \msb, \myTheta{C} is half of the upper
limit of the continuous uniform prior on the mean of the descendant population
sizes.

@@

For all of the analyses below, I use four summary statistics for each of
the pairs of populations:
$\pi$ \citep{Tajima1983}, $\theta_W$ \citep{Watterson1975}, $\pi_{net}$
\citep{Takahata1985}, and $SD(\pi-\theta_W)$ \citep{Tajima1989}.
Furthermore, in addition to model parameters, each sample \hpvector{}
also contains four statistics that summarize \divTimeMapVector:
the mean (\divTimeMean), variance (\divTimeVar), dispersion index
($\divTimeDispersion = \divTimeVar/\divTimeMean$), and the number
of divergence time parameters (\divTimeNum).
These have been denoted as \meant{}{}, \vart{}{}, \vmratio{}, and \numt{},
respectively \citep{Hickerson2006,Huang2011,Oaks2012}.
I use the new notation to avoid confusion that the values represent prior or
posterior expectations, and to indicate that \vmratio{} is not a parameter of
the model and \numt{} is no longer a parameter in the new implementation.

@@

The original descriptions of the \msb model claim that this re-ordering is
justified by the fact that of $\pi_b$ (and other summary statistics) are
unrelated to the sample sizes \popSampleSize{1}{} and \popSampleSize{2}{} of
each pair and are thus exchangeable \citep{Hickerson2006,Huang2011}.
This is actually incorrect for two reasons.
First, the expectation of $\pi_b$ is not independent of samples sizes.
If there are more than one coalescent events in the ancestor, which is expected
to be common on phylogeographic timescales, more samples will increase the
probability of capturing these deeper events, and thus affect the average
pairwise differences between the descendant populations.
Also, other statistics that estimate gross diversity (e.g., $\pi$ and
$\theta_W$) are clearly not independent of sample size.
Second, for variables to be exchangeable, they do not need to be independent,
but their marginal distributions must be the same (i.e., they must be
identically distributed).
The simulated alignments and their summary statistics are not identically
distributed because of differing sample sizes \emph{and} taxon- and
locus-specific parameters \hkyModel{}{}, \mutationRateScalarConstant{}{}, and
\ploidyScalar{}{}.

Thus, the theoretical basis for this reshuffling is questionable.
It may be justifiable as an additional coarsening of the data,
by removing the information associated with the identity of the
population pairs.
However, I do not provide a formal proof that this re-ordering of the pairs
does not introduce bias.
One can imagine situations in which the sampling intensity (i.e., the
number of gene copies, loci, and locus length) is highly skewed across
pairs and/or there is large heterogeneity in mutational processes
(e.g., \hkyModel{}{}, \mutationRateScalarConstant{}{}) among the pairs.
In such cases, it seems possible that discarding this information could
cause bias.
Furthermore, given that (1) part of the motivation for re-ordering by $\pi_b$
is to minimize Euclidean distances between simulated datasets in which the true
model has a single divergence event, and (2) the reordering increases the
model's tendency to infer a single divergence event \citep{Huang2011}, it seems
possible that this approach could be biasing the method \citep{Oaks2012}.

To maintain compatibility and comparability with \msb I maintain
the re-ordering of taxon-specific summary statistics by $\pi_b$.
However, I also allow the order to be maintained, and ordered
divergence models to be estimated.

\subsubsection{Obtaining an approximate posterior from the prior samples}
I use a rejection algorithm to retain an approximate posterior sample of
\paramSampleVector{} from the prior sample
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$.
First, the observed summary statistics \ssVectorObs, and the summary statistics
of the prior samples 
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$,
are standardized using the means and standard deviations of the statistics from
the prior sample (i.e., the prior mean is subtracted from each statistic, and the
difference is divided by the prior standard deviation).
After all statistics are standardized, the Euclidean distance between
\ssVectorObs and each \ssVector{} within \ssMatrix is calculated.
The samples that fall within a range of tolerance \tol around \ssVectorObs
are retained.
The range of tolerance is determined by specifying the desired number of
posterior samples to be retained.
Post-hoc adjustment of the posterior sample can also be performed with a number
of regression techniques \citep{Beaumont2002,Blum2009,Leuenberger2010}.
For my analyses, I use the general linear model (GLM) regression adjustment
\citep{Leuenberger2010} as implemented in \abctoolbox
\citep[v1.1;][]{ABCtoolbox}, which \citet{Oaks2012} showed performed very
similarly to weighted local-linear regression and multinomial logistic
regression adjustments \citep{Beaumont2002} for \msb posteriors.

\subsection{Assessing model-choice behavior and robustness}
Following the simulation-based approach of \citet{Oaks2012}, I characterize
the behavior of several models under the ideal conditions where the
data are generated from parameters drawn from the same prior distributions used
for analysis (i.e., the prior is correct).
I selected four model priors for these analyses (Table~\ref{tabPriors}):
(1) A model to represent the original \msb model, \modelOld, with
priors
$\divTimeIndexVector \sim \priorOld$,
$\divTime{} \sim U(0,10)$,
$\ancestralTheta{} \sim U(0, 0.05)$,
and
$\descendantThetaMean{} \sim U(0, 0.05)$;
(2) a model with the \msb prior on unordered divergence models, but with
exponential priors on nuisance parameters,
\modelUshaped, with priors
$\divTimeIndexVector \sim \priorOld$,
$\divTime{} \sim Exp(mean=2.887)$,
$\ancestralTheta{} \sim Exp(mean=0.025)$,
$\descendantTheta{1}{} \sim Exp(mean=0.025)$,
and
$\descendantTheta{2}{} \sim Exp(mean=0.025)$.
(3) a model with uniform prior probability over unordered divergence models,
\modelUniform, with priors
$\divTimeIndexVector \sim \priorUniform$,
$\divTime{} \sim Exp(mean=2.887)$,
$\ancestralTheta{} \sim Exp(mean=0.025)$,
$\descendantTheta{1}{} \sim Exp(mean=0.025)$,
and
$\descendantTheta{2}{} \sim Exp(mean=0.025)$.
and
(4) a Dirichlet-process prior model, \modelDPP, with priors
$\divTimeIndexVector \sim \priorDPP{\sim Gamma(2,2)}$,
$\divTime{} \sim Exp(mean=2.887)$,
$\ancestralTheta{} \sim Exp(mean=0.025)$,
$\descendantTheta{1}{} \sim Exp(mean=0.025)$,
and
$\descendantTheta{2}{} \sim Exp(mean=0.025)$.
I selected the exponential prior on divergence time used in models \modelDPP,
\modelUniform, and \modelUshaped to have the same variance as the uniform prior
in model \modelOld.
I selected the exponential prior on population size used in models \modelDPP,
\modelUniform, and \modelUniform to have the same mean as the uniform prior in
model \modelOld, so that all four models have the same \myTheta{C} and thus the
same units of time.
All of the models were the same in other respects, with three free \myTheta{}
parameters for each population pair, two uniformly distributed ($beta(1,1)$)
\bottleScalar{}{} parameters per pair, no migration, no recombination,
and unordered divergence models.
For all of my simulations, I used a simulated data structure of eight
population pairs, with a single 1000 base-pair locus sampled from 10
individuals from each population.

For each of the four models, I simulated $1\e6$ samples from the prior and
50,000 datasets, also drawn from the prior.
I then analyzed each of the simulated datasets, retaining a
posterior of 1000 samples from the respective prior.
A GLM-regression adjusted posterior was also estimated from each of the
posterior samples \citep{Leuenberger2010}.
To assess the robustness of each of the four models, I also analyzed the
datasets simulated under the other three models.
Overall, for each model, I produced 200,000 posterior estimates,
50,000 from the datasets simulated under that model,
and 150,000 estimated from the datasets simulated under the
other three models.

For each set of 50,000 simulated datasets, I used the posterior estimates
to assess the model-choice behavior of each model.
I did this by assigning the 50,000 estimates of the posterior probability
of one divergence event to 20 bins of width 0.05, and plotted
the estimated posterior probability of each bin against the proportion of
replicates in that bin with a true value consistent with one divergence
event \citep{Huelsenbeck2004,Oaks2012}.
I did this using two criteria for the one divergence model:
(1) the number of divergence time parameters ($\divTimeNum = 1$) and
(2) the dispersion index of divergence times \citep[$\divTimeDispersion <
0.01$][]{Hickerson2006,Huang2011}.
I used the one-divergence model to assess model-choice behavior, because
it is often of biogeographic interest and is easily comparable among
the three different priors used on divergence models.

In addition to the four models above, I also assessed the behavior of a model
with ordered divergence models (i.e., the order of the taxon-specific summary
statistic vectors were maintained for the observed and simulated datasets).
For this, I used a model with identical priors as \modelDPP, but that samples
over ordered divergence models.
I denote this model as \modelDPPOrdered.
I simulated $1\e6$ prior samples and 50,000 datasets, and
analyzed them as above.
I was not able to analyze the simulated datasets of the other models under
the ordered model, because the identity of the population pairs is not
contained in the simulations of the other models.

>>>>

\begin{linenomath}
Each \divTime{} within \divTimeVector is \iid as $\divTime{} \sim
Gamma(\gammaShape{\divTime{}}, \gammaScale{\divTime{}})$.
Thus, given the number of unique divergence-time classes in
\divTimeIndexVector, this determines the probability of prior term
$p(\divTimeVector \given \divTimeIndexVector)$.
The divergence times are in coalescent units relative to the size of
a constant reference population, which I denote \myTheta{C}, that is equal to
the expectation of the prior on the size of the descendant populations
\begin{equation}
    % \myTheta{C} = \mathbb{E}(\dThetas),
    \myTheta{C} = \mathbb{E}(\descendantTheta{}{}),
    \label{eq:thetaC}
\end{equation}
% where \popIndex is the descendant-population index which can be either 1 or 2.
Given the size of the descendant populations are \iid as
$\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$ and
$\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
this becomes
\begin{equation}
    \myTheta{C} = \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}.
    \label{eq:thetaCGamma}
\end{equation}
More specifically, the \divTime{} parameters are in units of
$\myTheta{C}/\mutationRate$ generations, which I denote as \globalcoalunit
generations.
Thus, each \divTime{} within \divTimeVector is proportional to time and can be
converted to the number of generations of the reference population, which I
denote \divTime{G_C}, by assuming a mutation rate and multiplying by the
effective size of the reference population
\begin{equation}
    \divTime{G_C} = \divTime \times \frac{\myTheta{C}}{\mutationRate} = \divTime
    \times \frac{
    \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}}
    {\mutationRate}.
    \label{eq:divTimeGenerations}
\end{equation}
Thus, for each of the divergence times in \divTimeVector to be on the same
scale, the relative mutation rates among the pairs of populations are assumed
to be known and fixed according to the user-provided values in
\mutationRateScalarConstantVector.
\end{linenomath}

@@

The second prior is based on the Dirichlet process, which is a stochastic
process that groups random variables into an unknown number of discrete
parameter classes \citep{Ferguson1973,Antoniak1974}.
The Dirichlet process has been used as a nonparametric Bayesian approach to
many inference problems in evolutionary biology
\citep{Lartillot2004,Huelsenbeck2007,HuelsenbeckSuchard2007,Ane2007,Heath2011,Heath2012}.
Here, I use the Dirichlet process to place a prior over all possible ordered
partitions of \npairs{} population pairs into divergence-time parameter classes
(i.e., ``divergence events'').
As discussed above, the time of each divergence-time parameter is drawn from
the base distribution $\divTime{} \sim Gamma(\gammaShape{\divTime{}},
\gammaScale{\divTime{}})$.
The partitioning of the population pairs to divergence-time classes is
controlled by the concentration parameter \concentrationParam, which determines
how clustered the process will be.
I take a hierarchical approach and use a prior probability
distribution (i.e., hyperprior) for \concentrationParam \citep{Escobar1995}.
More specifically, I use a gamma-distributed prior $\concentrationParam \sim
Gamma(\gammaShape{\concentrationParam}, \gammaScale{\concentrationParam})$,
where \gammaShape{\concentrationParam} and \gammaScale{\concentrationParam} are
specified by the investigator.
I use $\divTimeIndexVector \sim \priorDPP{}$ to denote this Dirichlet-process
prior.

@@

\subsection{Differences between this model and the original \texttt{msBayes} model}
\subsubsection{The prior on divergence models}
One of the key differences between my model and that of \msb \citep{Huang2011}
is the prior distribution on divergence models.
As discussed in \citet{Oaks2012}, in \msb the prior used for
\divTimeIndexVector is a combination of a discrete uniform prior over the
possible number of divergence events \divTimeNum from 1 to \npairs{} with a
multinomial distribution on the number of times each index of \divTimeVector
appears in \divTimeIndexVector, with the constraint that all \divTime{}
parameters are represented at least once (see Equation 2 of \citet{Oaks2012}).
I denote this prior used by \msb as $\divTimeIndexVector \sim \priorOld$.
\citet{Oaks2012} discuss how placing a uniform prior over the number of
divergence parameters (denoted \divTimeNum here, and as \numt{} in
\citet{Huang2011}) imposes an U-shaped prior over discrete divergence
models (\divTimeIndexVector; see
Figure 5B of \citet{Oaks2012}).
To avoid this, I place priors directly on the sample space of divergence
models, thus eliminating the parameter \numt{} from the model.
I introduce two priors on divergence models:
(1) a prior that is uniform over all unordered divergence models, and
(2) a Dirichlet-process prior on all ordered divergence models.
The latter provides an investigator with a great deal of flexibility in
expressing their prior beliefs about models of divergence.

@@

\subsubsection{Flexibility in parameterizing the model}
In the new implementation, I provide the ability to control the richness of the
model.
For the \myTheta{} parameters, the model can be fully generalized to
allow each population pair to have three parameters:
\ancestralTheta{}, \descendantTheta{1}{}, and \descendantTheta{2}{}.
Furthermore, any model of \myTheta{} parameters nested within this
general model can also be specified, including the most restricted model
where the ancestral and descendant populations of each pair share
a single \myTheta{} parameter.

@@

Overall, my implementation allows an investigator to specify a model that has
as many as seven parameters per population pair
(\ancestralTheta{}, \descendantTheta{1}{}, \descendantTheta{2}{},
\bottleTime{}, \bottleScalar{1}{}, \bottleScalar{2}{}, and
\migrationRate{})
or as few as one parameter per pair
(\myTheta{}),
in addition to the $\popSampleSize{i}{} - 1$ coalescence-time parameters
(i.e., the node heights of the gene tree).

\subsubsection{Time scale}
As described above, divergence times are in units of
$\myTheta{C}/\mutationRate$ generations, where \myTheta{C} is the expectation
of the prior on descendant-population size.
As described by \citet{Oaks2012}, in \msb, \myTheta{C} is half of the upper
limit of the continuous uniform prior on the mean of the descendant population
sizes.
This is only equal to the expectation of the prior if the lower limit is zero.

@@

For all of the analyses below, I use four summary statistics for each of
the pairs of populations:
$\pi$ \citep{Tajima1983}, $\theta_W$ \citep{Watterson1975}, $\pi_{net}$
\citep{Takahata1985}, and $SD(\pi-\theta_W)$ \citep{Tajima1989}.
Furthermore, in addition to model parameters, each sample \hpvector{}
also contains four statistics that summarize \divTimeMapVector:
the mean (\divTimeMean), variance (\divTimeVar), dispersion index
($\divTimeDispersion = \divTimeVar/\divTimeMean$), and the number
of divergence time parameters (\divTimeNum).
Previously, these have been denoted as \meant{}{}, \vart{}{}, \vmratio{}, and
\numt{}, respectively \citep{Hickerson2006,Huang2011,Oaks2012}.
I use \divTimeMean and \divTimeVar in place \meant{}{} and \vart{}{} to make
clear that these values do not represent the prior or posterior
expectation/variance of divergence times.
I use \divTimeDispersion in place of \vmratio{} to clarify that this is a
statistic rather than a parameter of the model.
Lastly, I use \divTimeNum in place of \numt{}, because the number of
divergence-time parameters is no longer a parameter in the new implementation.

@@

The original descriptions of the \msb model claim that this re-ordering is
justified by the fact that the expected value of $\pi_b$ is
unrelated to sample size \popSampleSize{1}{} and \popSampleSize{2}{} and
thus exchangeable among pairs \citep{Hickerson2006,Huang2011}.
This is actually incorrect for three reasons.
First, the independence of the expected value of $\pi_b$ and sample size
breaks down as $\pi_b$ approaches its lower limit of zero (i.e., very recent or
no divergence between populations), which might be common given the timescales
of most phylogeographical studies.
Second, the entire vector of summary statistics \alignmentSS{}{} for each pair
of populations is re-ordered across pairs, which implies that the justification
for re-ordering $\pi_b$ applies to all the statistics within each
\alignmentSS{}{}.
% First, the expectation of $\pi_b$ is not independent of samples sizes.
% If there are more than one coalescent events in the ancestor, which is expected
% to be common on phylogeographic timescales, more samples will increase the
% probability of capturing these deeper events, and thus affect the average
% pairwise differences between the descendant populations.
However, the expectations for statistics that estimate gross diversity (e.g.,
$\pi$ and $\theta_W$) are not independent of sample size for structured
populations, and other statistics are not independent of sample size in general
(e.g., $SD(\pi-\theta_W)$).
Third, for variables to be exchangeable, they do not need to be independent,
but their marginal distributions must be the same (i.e., they must be
identically distributed).
The simulated alignments and their summary statistics are not identically
distributed because of differing sample sizes \emph{and} taxon- and
locus-specific parameters \hkyModel{}{}, \mutationRateScalarConstant{}{}, and
\ploidyScalar{}{}.

Thus, the theoretical basis for this reshuffling is questionable.
It may be justifiable as an additional coarsening of the data,
by removing the information associated with the identity of the
population pairs.
However, I do not provide a formal proof that this re-ordering of the pairs
does not introduce bias.
One can imagine situations in which the sampling intensity (i.e., the
number of gene copies, loci, and locus length) is highly skewed across
pairs and/or there is large heterogeneity in mutational processes
(e.g., \hkyModel{}{}, \mutationRateScalarConstant{}{}) among the pairs.
In such cases, it seems possible that discarding this information could
cause bias.
Furthermore, given that (1) part of the motivation for re-ordering by $\pi_b$
is to minimize Euclidean distances between simulated datasets in which the true
model has a single divergence event \citep{Huang2011}, and (2) the reordering
increases the model's tendency to infer a single divergence event
\citep{Huang2011}, it seems possible that this approach could be biasing the
method \citep{Oaks2012}.

For comparability with \msb, I maintain the re-ordering of taxon-specific
summary statistics by $\pi_b$.  However, I also allow the order to be
maintained, and ordered divergence models to be estimated.

\subsubsection{Obtaining an approximate posterior from the prior samples}
I use a rejection algorithm to retain an approximate posterior sample of
\paramSampleVector{} from the prior sample
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$.
First, the observed summary statistics \ssVectorObs, and the summary statistics
of the prior samples 
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$,
are standardized using the means and standard deviations of the statistics from
the prior sample (i.e., the prior mean is subtracted from each statistic, and the
difference is divided by the prior standard deviation).
After all statistics are standardized, the Euclidean distance between
\ssVectorObs and each \ssVector{} within \ssMatrix is calculated.
The samples that fall within a range of tolerance \tol around \ssVectorObs
are retained.
The range of tolerance is determined by specifying the desired number of
posterior samples to be retained.
Post-hoc adjustment of the posterior sample can also be performed with a number
of regression techniques \citep{Beaumont2002,Blum2009,Leuenberger2010}.
For my analyses, I use the general linear model (GLM) regression adjustment
\citep{Leuenberger2010} as implemented in \abctoolbox
\citep[v1.1;][]{ABCtoolbox}, which \citet{Oaks2012} showed performs very
similarly to weighted local-linear regression and multinomial logistic
regression adjustments \citep{Beaumont2002} for \msb posteriors.

\subsection{Assessing model-choice behavior and robustness}
Following the simulation-based approach of \citet{Oaks2012}, I characterize
the behavior of several models under the ideal conditions where the
data are generated from parameters drawn from the same prior distributions used
for analysis (i.e., the prior is correct).
I selected four model priors for these analyses (Table~\ref{tabPriors}):
\begin{description}
    \item[\modelOld] a model to represent the original \msb implementation with
        the U-shaped prior on unordered divergence models and uniform priors on
        divergence-time and demographic parameters; $\divTimeIndexVector \sim
        \priorOld$, $\divTime{} \sim U(0,10)$, $\ancestralTheta{} \sim U(0,
        0.05)$, and $\descendantThetaMean{} \sim U(0, 0.05)$
    \item[\modelUshaped] a model with the U-shaped prior of \msb on unordered
        divergence models, but with exponential priors on divergence-time and
        demographic parameters; $\divTimeIndexVector \sim \priorOld$,
        $\divTime{} \sim Exp(mean=2.887)$, $\ancestralTheta{} \sim
        Exp(mean=0.025)$, $\descendantTheta{1}{} \sim Exp(mean=0.025)$, and
        $\descendantTheta{2}{} \sim Exp(mean=0.025)$
    \item[\modelUniform] a model with a uniform prior over unordered divergence
        models and exponential priors on divergence-time and demographic
        parameters; $\divTimeIndexVector \sim \priorUniform$, $\divTime{} \sim
        Exp(mean=2.887)$, $\ancestralTheta{} \sim Exp(mean=0.025)$,
        $\descendantTheta{1}{} \sim Exp(mean=0.025)$, and
        $\descendantTheta{2}{} \sim Exp(mean=0.025)$
    \item[\modelDPP] a model with a Dirichlet-process prior on ordered
        divergence models and exponential priors on divergence-time and
        demographic parameters; $\divTimeIndexVector \sim \priorDPP{\sim
        Gamma(2,2)}$, $\divTime{} \sim Exp(mean=2.887)$, $\ancestralTheta{}
        \sim Exp(mean=0.025)$, $\descendantTheta{1}{} \sim Exp(mean=0.025)$,
        and $\descendantTheta{2}{} \sim Exp(mean=0.025)$.
\end{description}
I selected the exponential prior on divergence time used in models \modelDPP,
\modelUniform, and \modelUshaped to have the same variance as the uniform prior
in model \modelOld.
I selected the exponential prior on population size used in models \modelDPP,
\modelUniform, and \modelUniform to have the same mean as the uniform prior in
model \modelOld, so that all four models have the same \myTheta{C} and thus the
same units of time.
All of the models were the same in other respects, with three free \myTheta{}
parameters for each population pair, two uniformly distributed ($beta(1,1)$)
\bottleScalar{}{} parameters per pair, no migration, no recombination, and
resorting of taxon-specific summary statistics by $\pi_b$ (i.e., sampling
unordered divergence models).
For all simulations, I used a data structure of eight population pairs, with a
single 1000 base-pair locus sampled from 10 individuals from each population.

For each of the four models, I simulated $1\e6$ samples from the prior and
50,000 datasets, also drawn from the prior.
I then analyzed each of the simulated datasets, retaining a
posterior of 1000 samples from the respective prior.
A GLM-regression adjusted posterior was also estimated from each of the
posterior samples \citep{Leuenberger2010}.
To assess the robustness of each of the four models, I also analyzed the
datasets simulated under the other three models.
Overall, for each model, I produced 200,000 posterior estimates,
50,000 from the datasets simulated under that model,
and 150,000 from the datasets simulated under the
other three models.

For each set of 50,000 simulated datasets, I used the posterior estimates
to assess the model-choice behavior of each model.
I did this by assigning the 50,000 estimates of the posterior probability
of one divergence event to 20 bins of width 0.05, and plotted
the estimated posterior probability of each bin against the proportion of
replicates in that bin with a true value consistent with one divergence
event \citep{Huelsenbeck2004,Oaks2012}.
I did this using two criteria for the one divergence model:
(1) the number of divergence time parameters ($\divTimeNum = 1$) and
(2) the dispersion index of divergence times \citep[$\divTimeDispersion <
0.01$;][]{Hickerson2006,Huang2011}.
I used the one-divergence model to assess model-choice behavior, because
it is often of biogeographic interest and is easily comparable among
the three different priors used on divergence models.

In addition to the four models above, I also assessed the behavior of a model
that samples over ordered divergence models (i.e., the order of the
taxon-specific summary statistic vectors were maintained for the observed and
simulated datasets); all other settings were identical to the \modelDPP model.
I denote this model as \modelDPPOrdered.
I simulated $1\e6$ prior samples and 50,000 datasets, and
analyzed them as above.
I was not able to analyze the simulated datasets of the other models under
the ordered model, because the identity of the population pairs is not
contained in the simulations of the other models.
